---
title: "Presenting Results: Visualizing Simulation Outcomes"
author: "Ashley I Naimi"
date: "`r paste0('Spring ', 2024)`" #format(Sys.Date(), '%Y')
urlcolor: blue
bibliography: ref.bib
link-citations: yes
output: 
    bookdown::pdf_book:
      base_format: tint::tintPdf
      toc: true
      number_sections: true
      includes:
        in_header: ../../misc/preamble.tex
      latex_engine: xelatex
    html_document:
      theme: readable
      toc: true
      toc_float: true
      number_sections: true
      css: ../../misc/style.css
---

```{r setup, include=FALSE}
library(knitr)
library(formatR)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

packages <- c( "data.table","tidyverse","ggplot2","ggExtra","formatR",
               "gridExtra","skimr","here","RColorBrewer","survival")

for (package in packages) {
  if (!require(package, character.only=T, quietly=T)) {
    install.packages(package, repos='http://lib.stat.cmu.edu/R/CRAN')
  }
}

for (package in packages) {
  library(package, character.only=T)
}

remotes::install_github("rstudio/fontawesome")

library(fontawesome)

thm <- theme_classic() +
  theme(
    legend.position = "top",
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )
theme_set(thm)

knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_chunk$set(echo = TRUE)

```

\newpage
\onehalfspacing

# Visualizing Simulation Outcomes

Once a simulation study is complete, the outcome of the study often consists of several datasets that contain information on the performance of the methods being studied. These datasets must then be processed in order to present the findings from the study. 

Several different forms of presentation can be used, including tables and figures. Besides the common repertoire of figures that can be used to present results (histograms, density plots, scatter plots, heat maps and other), there are a few types of figures that can be tailored to a simulation study, and that can potentially reveal useful information about the performance of a set of methods under a particular set of circumstances. 

Here, we will briefly introduce nested loop plots, and zipper (or "zip")^[Sometimes zipper plots are referred to as zip plots [e.g., @Morris2019]. However, searching the term "zip plot" online will yield many plot structures tailored to plotting geographical regions using zip codes.] plots, and demonstrate how they can be used to convey results from a simulation study.

# Nested Loop Plot

Nested loop plots can be used to represent the results of a simulation study over all possible cross-combination of parameters used to define the simulation data.^[this section was based heavily on a very useful site by Michael Kammer: https://bit.ly/4bm77eP \\ Thanks to an Emory Epi PhD student, Qi Zhang, for bringing this to my attention.]

Suppose we conduct a simulation study evaluating the performance of three different methods under a range of difference scenarios. Suppose further that from our simulation code, we obtain a dataset with the following information:

```{r tidy = F, warning = F, message = F}

pacman::p_load(
  tidyverse,     
  dplyr, 
  purr,
  magrittr
  )

thm <- theme_classic() +
  theme(
    legend.position = "top",
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )
theme_set(thm)

set.seed(123)
params = list(
    samplesize = c(100, 200, 500),
    param1 = c(1, 2), 
    param2 = c(1, 2, 3), 
    param3 = c(1, 2, 3, 4)
)

design = expand.grid(params)

# add some "results"
design %<>% 
    mutate(method1 = rnorm(n = n(),
                           mean = param1 * (param2 * param3 + 1000 / samplesize), 
                           sd = 2), 
           method2 = rnorm(n = n(),
                           mean = param1 * (param2 + param3 + 2000 / samplesize), 
                           sd = 2), 
           method3 = rnorm(n = n(),
                           mean = param1 * (param2 + param3 + 3000 / samplesize), 
                           sd = 2))

knitr::kable(head(design, n = 10))
```

We can use a nested loop plot to present these results in a single image. To deploy a nested loop plot, we first need to install the relevant package. This package is a development package hosted on GitHub, so we'll need to use the `remotes` or `devtools` packages to install it:

```{r tidy = F, warning = F, message = F, results='hide'}

remotes::install_github("matherealize/looplot")

```


We can then proceed to use the `nested_loop_plot` function, which relies on `ggplot2` functionality. With the dataset above, we can construct a plot using the following code:

```{r tidy = F, warning = F, message = F}

pacman::p_load(looplot)

p = nested_loop_plot(resdf = design, 
                     x = "samplesize", steps = c("param2", "param3"),
                     grid_rows = "param1", 
                     steps_y_base = -10, steps_y_height = 3, steps_y_shift = 10,
                     x_name = "Sample Size", y_name = "Error",
                     spu_x_shift = 200,
                     steps_values_annotate = TRUE, steps_annotation_size = 3, 
                     hline_intercept = 0, 
                     y_expand_add = c(10, NULL), 
                     post_processing = list(
                        add_custom_theme = list(
                            axis.text.x = element_text(angle = -90,
                                                       vjust = 0.5,
                                                       size = 8)
                        ))
                     )


ggsave(here("_images", "nested_loop_plot.pdf"), width = 8, height = 6)
```

Here is what the figure we generated looks like:

```{r nestedloop1, out.width="12cm", fig.align='center', fig.cap="Example Nested Loop Plot of Hypothetical Simulation Results.", echo=F}
knitr::include_graphics(here("_images", "nested_loop_plot.pdf"))
```

This Figure shows the magnitude of the simulated error for each sample size split by the two distinct `param1` values, across all combinations of `param2` and `param3`.

There are many different ways to formulate a plot like this. For example, we can remove the separate across `param1`

```{r tidy = F, warning = F, message = F}

pacman::p_load(looplot)

p = nested_loop_plot(resdf = design, 
                     x = "samplesize", steps = c("param1", "param2", "param3"),
                     #grid_rows = "param1", 
                     steps_y_base = -10, steps_y_height = 3, steps_y_shift = 10,
                     x_name = "Sample Size", y_name = "Error",
                     spu_x_shift = 200,
                     steps_values_annotate = TRUE, steps_annotation_size = 3, 
                     hline_intercept = 0, 
                     y_expand_add = c(10, NULL), 
                     post_processing = list(
                        add_custom_theme = list(
                            axis.text.x = element_text(angle = -90,
                                                       vjust = 0.5,
                                                       size = 8)
                        ))
                     )


ggsave(here("_images", "nested_loop_plot2.pdf"), width = 10, height = 6)
```

Which gives us a single panel figure:

```{r nestedloop2, out.width="12cm", fig.align='center', echo=F}
knitr::include_graphics(here("_images", "nested_loop_plot2.pdf"))
```

We can add separate grids for each parameter as well. For a range of different options of the nested loop plot, refer to the package demo: https://bit.ly/4bm77eP.

# Zipper Plots

Zipper plots are most often used to present bounds, such as confidence intervals. Consider data from the `rsimsum` package on the performance of different methods to estimate a hazard ratio when the baseline hazard is misspecified:

```{r tidy = F, warning = F, message = F, echo = F}
pacman::p_load(rsimsum)

data("relhaz", package = "rsimsum")

head(relhaz)

dim(relhaz)

table(relhaz$n)
table(relhaz$baseline)
table(relhaz$model)

```

The survival outcomes in each dataset were simulated from a binary treatment variable with a log-hazard ratio of -0.50, under sample sizes of 50 and 250 individuals, and under two different baseline hazard functions (exponential and Weibull). We can also see that for each combination of simulation parameters (sample size, baseline hazard, and model), the Monte Carlo sample size was 100:

```{r tidy = F, warning = F, message = F}

relhaz %>% 
  group_by(n, baseline, model) %>% 
  count()

```


Each of the 100 simulated datasets was then analyzed using a Cox proportional hazards regression model, a parametric exponential model, and a flexible parametric model developed by Patric Royston and Mahesh Parmar, where the baseline hazard is fit with natural cubic splines with two degrees of freedom [@Royston2002].^[Additional details on this dataset are available here: https://bit.ly/3K6zpOr]

Suppose we're interested in exploring the performance of the normal-interval (or Wald) confidence interval estimator in these data. We can construct upper and lower confidence interval bounds in the data above using the standard equation:

$$(LCL, UCL) = \hat{\theta} \pm 1.96 \times SE(\hat{\theta})$$

In R, we could implement this as follows:

```{r tidy = F, warning = F, message = F}

relhaz <- relhaz %>% 
  mutate(lcl = theta - 1.96*se,
         ucl = theta + 1.96*se)

head(relhaz)

```

We can also add an indicator of whether the bounds just created include the true value of -0.5 or not:

```{r tidy = F, warning = F, message = F}
relhaz <- relhaz %>% 
  mutate(include_flag = if_else(lcl<-.5 & ucl>-.5, "Include", "Exclude"))
```

With these upper and lower bounds, we can now create a zipper plot for a subset of these results:

```{r tidy = F, warning = F, message = F}

p <- relhaz %>% 
  filter(n == 50, baseline == "Exponential") %>% 
  ggplot(.) + 
  geom_hline(yintercept = -.5, lty = 2) +
  geom_pointrange(aes(x = dataset, 
                      y = theta, 
                      ymin = lcl, 
                      ymax = ucl, color = include_flag), 
                  size = .2, 
                  alpha = .75) + 
  scale_color_manual(values=c("red","grey")) +
  ylab("log Hazard Ratio") + 
  xlab("Sample Number") +
  coord_flip() +
  theme(legend.position = "none", text=element_text(size=12)) +
  facet_wrap(~model)

ggsave(here("_images", "zip_plot_version1.pdf"), p)

```

This plot reveals the distribution of confidence intervals across all 100 iterations:

```{r zipper1, out.width="10cm", fig.align='center', fig.cap="Zipper plot displaying the distribution of normal-interval (Wald) confidence intervals in the relhaz data.", echo=F}
knitr::include_graphics(here("_images", "zip_plot_version1.pdf"))
```

Some authors like to present these zipper plots with the bounds ranked according to some criterion. For confidence intervals, we can rank our results according to the magnitude of the Wald test statistic for, say, a null test hypothesis:

```{r tidy = F, warning = F, message = F}

relhaz <- relhaz %>% 
  mutate(test_statistic = abs(theta/se))

```

We can then incorporate an `arrange` argument into our plot code:

```{r tidy = F, warning = F, message = F}

p <- relhaz %>% 
  filter(n == 50, baseline == "Exponential") %>% 
  ggplot(.) + 
  geom_hline(yintercept = -.5, lty = 2) +
  geom_pointrange(aes(x = test_statistic, 
                      y = theta, 
                      ymin = lcl, 
                      ymax = ucl, color = include_flag), 
                  size = .2, 
                  alpha = .75) + 
  scale_color_manual(values=c("red","grey")) +
  ylab("log Hazard Ratio") + 
  xlab("Wald Null Test Statistic") +
  coord_flip() +
  theme(legend.position = "none", text=element_text(size=12)) +
  facet_wrap(~model)

ggsave(here("_images", "zip_plot_version2.pdf"), p)

```

Which gives us the following modified figure:

```{r zipper2, out.width="10cm", fig.align='center', fig.cap="Zipper plot displaying the distribution of normal-interval (Wald) confidence intervals in the relhaz data. Bounds are ranked according to the magnitude of the Wald test statistic for each point estimate.", echo=F}
knitr::include_graphics(here("_images", "zip_plot_version2.pdf"))
```

Alternative rankings for the y-axis can be considered, such as the magnitude of the standard error, or the centile rank of the test statistic. 

\newpage

# References