---
title: "Simulating Random Variables in R"
author: "Ashley I Naimi"
date: "Summer `r format(Sys.Date(), '%Y')`"
urlcolor: blue
bibliography: ../misc/ref.bib
link-citations: yes
output: 
    bookdown::pdf_book:
      base_format: tint::tintPdf
      toc: true
      number_sections: true
      includes:
        in_header: ../misc/preamble.tex
      latex_engine: xelatex
    html_document:
      theme: readable
      toc: true
      toc_float: true
      number_sections: true
      css: ../misc/style.css
---

```{r setup, include=FALSE}

# Load pacman (install if necessary)
if (!requireNamespace("pacman", quietly = TRUE)) {
  install.packages("pacman", repos = 'http://lib.stat.cmu.edu/R/CRAN')
}

# Use pacman to load (and install if missing) all required packages
pacman::p_load(
  data.table, tidyverse, ggplot2, ggExtra, formatR,
  gridExtra, skimr, here, RColorBrewer,
  knitr, remotes
)

# Ensure fontawesome is installed from GitHub if not available
if (!requireNamespace("fontawesome", quietly = TRUE)) {
  remotes::install_github("rstudio/fontawesome")
}
library(fontawesome)

# Set ggplot2 theme
thm <- theme_classic() +
  theme(
    legend.position = "top",
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )
theme_set(thm)

# Set global chunk options
opts_chunk$set(
  tidy.opts = list(width.cutoff = 60),
  tidy = TRUE,
  echo = TRUE
)

# Set knit hook for purling
knit_hooks$set(purl = hook_purl)

```

\newpage
\onehalfspacing

# Distribution Functions in R

Base R comes with a wide variety of pseudo-random number generators that we can use to simulate from a wide variety of probability distributions, including the Gaussian, uniform, binomial, Poisson, and other distributions. Additionally, there are several packages written that can be used to simulate from other common and less common distributions, including the multivariate Normal, double-exponential, Gumbel, and others. Generally, functions written in R to generate data from a distribution follow a particular convention. For example, generating from a normal distribution, we have the following functions in base R:

  - rnorm: generate Normal random variable
  - dnorm: evaluate the Normal probability density (with a given mean/SD) at specific points
  - pnorm: evaluate the cumulative distribution function for a Normal distribution 
  - qnorm: evaluate the inverse of the cumulative distribution function for a Normal distribution 
  
Generally, functions that start with an "r" generate random variables. Functions that start with a "d" generate density values for a specific value of the random variable. Functions that start with a "p" and a "q" evaluate the cumulative distribution function, and the inverse of the cumulative distribution (quantile) function, respectively.

|  Prefix  | Function |
|----|---|
| d  |  density |
| r  |  random variable generation |
| p  |  cumulative distribution |
| q  |  inverse cumulative distribution (quantile) |


Mostly, to conduct simulation studies, we'll rely on the "r" functions, which will allow us to generate random variables. Next we'll look at specific functions that generate random variables in R and explore their use for simulation.

## Gaussian (or Normal) Distribution

We can simulate data from a Normal distribution using the `rnorm` function in R. This function can be deployed as:

```{r gaussianplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Histogram for Univariate Normal Distribution with Mean = 0 and Standard Deviation = 1 for 5000 Simulated Observations."}

set.seed(123)
ggplot() + 
  geom_histogram(aes(x = rnorm(n = 5000))) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

```{r, warning = F, message = F}

set.seed(123)

n <- 5

y <- rnorm(n, mean = 0, sd = 1)

y

```

The univariate Normal distribution is fully defined by its mean and standard deviation. The default values for these in R are 0 and 1, respectively. 

In a regression context, the mean of a Normally distributed random variable is often what is made conditional on other variables. For example, in a simple setting, we might have a Normally distributed outcome $Y$ with a mean conditional on $X$. This can be accomplished in R in two ways^[Note the use of three `set.seed()` functions in this code chunk. This is not advisable, but I'm doing it here to demonstrate an equivalence between two ways of generating a $Y$ variable conditional on $X$.]:

```{r, warning = F, message = F}

n <- 5

set.seed(123)
x <- rnorm(n)

set.seed(123)
y_version1 <- rnorm(n, mean = 1 + 2*x, sd = 1)

set.seed(123)
y_version2 <- 1 + 2*x + rnorm(n, mean = 0, sd = 1)

y_version1

y_version2

```

## Multivariate Normal Distribution

```{r mvnplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Contour Plot for Multivariate Normal Distribution with Mean = (0,0), Standard Deviation = (1,1), and covariance (.5, .5) for 5000 Simulated Observations."}

set.seed(123)
m <- c(0, 0)
sigma <- matrix(c(1,.5,.5,1), nrow=2)
data.grid <- expand.grid(x = seq(-3, 3, length.out=200), y = seq(-3, 3, length.out=200))
q.samp <- cbind(data.grid, prob = mvtnorm::dmvnorm(data.grid, mean = m, sigma = sigma))
ggplot(q.samp, aes(x=x, y=y, z=prob)) + 
    geom_contour() +
    coord_fixed(xlim = c(-3, 3), ylim = c(-3, 3), ratio = 1) 

```

It is sometimes useful to generate data from a multivariate Normal distribution. For example, if you are trying to simulate a large number of independent covariates (e.g., confounders or predictors) that will be included in a regression model, rather than copy and paste code for the univariate Normal distribution multiple times, you can use code for generating a multivariate Normal vector, and make the covariance between them zero. 

Alternatively, if you are interested in exploring the impact of non-zero covariance between a set of covariates on the performance of an estimator, you can use multivariate Normal functions to do so. 

There are two packages in R that can be used to generate multivariate Normal data: the `MASS` package and the `mvtnorm` package. Here's how to use the functions in `MASS`:

```{r, warning = F, message = F}

  n <- 5

  set.seed(123)

  # create variance - covariance matrix:
  sigma <- matrix(0,nrow=3,ncol=3) 
  diag(sigma) <- 1
  
  sigma
  
  # create mean vector:
  mu <- rep(1, 3)
  
  mu
   
  # simulate variables
  c <- MASS::mvrnorm(n, mu=mu, Sigma=sigma)
  
  c

```

Here's how to do the same thing in `mvtnorm`:

```{r, warning = F, message = F}

  n <- 5

  set.seed(123)

  # create variance - covariance matrix:
  sigma <- matrix(0,nrow=3,ncol=3) 
  diag(sigma) <- 1
  
  sigma
  
  # create mean vector:
  mu <- rep(1, 3)
  
  mu
   
  # simulate variables
  c <- mvtnorm::rmvnorm(n, mean=mu, sigma=sigma)
  
  c

```

Note the flexibility that can be introduced into these function calls. You can specify different means for each column, different standard deviations for each column, as well as different variance-covariance relations between columns, all depending on the nature of the question you are interested in answering. 

:::{.rmdnote data-latex="{note}"}

__Technical Note__:

|               Recently in our work, we have been exploring the impact of increasing or decreasing the number of variables in a regression model on the performance of a range of estimators. There is an easy way to do this using matrix multiplication in R. To do this, we are using the `mvtnorm` package, with code that looks like this:

```{r  tidy = F, warning = F, message = F}
  
library(mvtnorm)

n <- 5
p <- c_number <- 3

# Covariate matrix
sigma <- diag(p)
c <- rmvnorm(n, mean = rep(0, p), sigma = sigma)

# Outcome model
parmsC <- rep(1.5, c_number)
y <- 10 + c %*% parmsC + rnorm(n)

# Final data
data <- data.frame(y, c)

```

The portion of code that multiplies the covariates by the parameters, `c %*% parmsC` represents a matrix multiplication operation, and can be used without the need to manually include additive terms. By simply changing the `c_number` value, you can automatically increase or decrease the number of covariates included in the model for $Y$.

This approach works for continuous variables entered into the outcome model for `y` linearly. It can also work for binary indicator variables. However, this code won't work for categorical variables without some additional pre-processing. An easy way to do this pre-processing is to rely on the `model.matrix()` function. 

:::

\newpage

## Uniform Distribution

```{r uniformplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Histogram for the Uniform Distribution with Upper and Lower Bounds of 0 and 1, respectively for 5000 Simulated Observations."}

set.seed(123)
ggplot() + 
  geom_histogram(aes(x = runif(n = 5000))) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

The uniform distribution is fully defined by its upper and lower bounds. The default values for these in R are 0 and 1, respectively. 

This distribution can be useful in a number of ways. In principle, this distribution can be made conditional on other variables by specifying the bounds of the distribution as a function of some other variable. However, this is not often seen in practice. One can generate a uniform random variable in R using the following code:

```{r}

set.seed(123)

n <- 5

y <- runif(n, min = 0, max = 1)

y

```

This distribution is central to the inverse transformation method, which is a generic technique used to simulate data from arbitrary distributions. We will see this method shortly.

## Binomial Distribution

```{r binomplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Barplot for the Binomial (Bernoulli) Distribution with p = 0.25 for 5000 Simulated Observations."}

set.seed(123)
ggplot() + 
  geom_bar(aes(x = rbinom(n = 5000, size = 1, p = .25))) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

The binomial distribution is defined by two parameters: the probability of a success in a given "trial" and the number of trial conducted. When the number of "trials" is one, the binomial distribution is equivalent to the Bernoulli distribution, which generates a $[0,1]$ indicator of whether an event occurred or not. In settings where logistic regression is used, the dependent variable is usually from a Bernoulli distribution with a probability depending on variables that will be included in the model. There are no default values for probability of success and the trial size, so these must be specified as follows. 

```{r}

set.seed(123)

n <- 5

y <- rbinom(n, size = 1, p = .5)

y

```

Note what happens when we increase the size of the trial to, say, 8:

```{r}

set.seed(123)

n <- 5

y <- rbinom(n, size = 8, p = .5)

y

```

Each instance of the simulated `y` becomes a sum of all of the successes (one's) encountered over the eight trials.

It's common to use a Bernoulli random variable in the context of a logistic regression model. We'll see how to do this in more depth later, but for now, note that we can define a linear term for a logistic regression model, define the inverse logistic link function, and then use these to generate a Bernoulli outcome $Y$ that depends on a single covariate $C$ through the link function:

```{r tidy = F, warning = F, message = F}

set.seed(123)

n <- 5

c <- rnorm(n)

expit <- function(x){1/(1+exp(-x))}

y <- rbinom(n, size = 1, p = expit(-2 + log(2)*c))

data <- data.frame(y, c)

data

```

It would help us if we could see that the relationship between $C$ and $Y$ here aligned with our logistic regression model, but we only generated five observations. In subsequent sections, we'll see similar code in action with much larger datasets.

## Multinomial Distribution

```{r multinomplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Barplot for the Multinomial Distribution with Three Levels and p = {0.2, 0.1, 0.7} for 5000 Simulated Observations."}

mn_vars <- t(rmultinom(n = 5000, size = 1, p = c(.2, .1, .7)))
mn_vars <- do.call(rbind, 
  lapply(1:nrow(mn_vars), function(x) which(mn_vars[x,]==1))
)

set.seed(123)
ggplot() + 
  geom_bar(aes(x = mn_vars)) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

The multinomial distribution is a generalization of the binomial distribution that can be used to generate categorical variables where the probability of each level is governed by a (possibly) unique probability value. The multinomial distribution is defined by two parameters: the set of values defining the probabilities of realizing a specific category. and the number of trial conducted. Similar to the binomial distribution, when the number of "trials" is one, the multinomial distribution is equivalent to the categorical distribution, which generates a $[0,1]$ indicator of whether an event occurred or not. In the multinomial (categorical) case, the "event" can take on more than 2 levels. There are no default values for probabilities of success and the trial size, so these must be specified. 

We can start by using the multinomial distribution in R to model the physical mechanism by which we roll a single (size = 1), six sided die. The "six-sidedness" comes from the fact we are repeating the "1/6" value six times (using the `rep()` function). Here, we roll a six-sided dice 5 times:

```{r}

set.seed(123)

n <- 5

y <- rmultinom(n, size = 1, p = rep(1/6, 6))

y

```

Alternatively, we can imagine a situation where an individual can experience one of three outcomes in a study related to cardiovascular health. Suppose the individual can be censored (y = 1), can experience a competing event such as death (y = 2), or can experience the outcome of interest, such as a heart attack (y = 3). Suppose further that these probabilities are 0.1, 0.05, and 0.15, respectively. This means that thirty percent of the sample experienced some event, and the remaining seventy percent experienced no event (such individuals would often be identified as administratively censored). We can simulate these data with the following code:

```{r tidy = F, warning = F, message = F}

set.seed(123)

n <- 5

y <- rmultinom(n, size = 1, 
               p = c(0.1,0.05,0.15, 1 - sum(0.1,0.05,0.15)))

y

t(y)

```

Note here that in this simulation, the first two observations experienced a competing event, the third and the fifth observations were censored, and the fourth observation experienced a heart attack.

<!-- ## Hypergeometric Distribution -->

<!-- ## Chi squared Distribution -->

## Poisson Distribution

```{r poisplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Histogram for the Poisson Distribution with lamba = 5 for 5000 Simulated Observations."}

set.seed(123)
ggplot() + 
  geom_bar(aes(x = rpois(n = 5000, lambda = 5))) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

The Poisson distribution is fully defined by a single parameter, usually denoted $\lambda$. This distribution is usually used to model counts of discrete events, such as the number of cigarettes smoked in one hour among smokers. A Poisson random variable can be simulated using the following code:

```{r}

set.seed(123)

n <- 5

y <- rpois(n, lambda = 3)

y

```

## Negative Binomial Distribution

```{r nbinomplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Histogram for the Negative Binomial Distribution with p = 0.5 and size = 100 for 5000 Simulated Observations."}

set.seed(123)
ggplot() +
  geom_bar(aes(x = rnbinom(n = 5000, prob = .2, size = 10) )) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

The negative binomial distribution is commonly used in epidemiology. There are a few ways in which one can represent a negative binomial distribution. The first is in relation to the standard binomial. This representation is sometimes referred to as the Pascal distribution [e.g., @Leemis2008] or the "counting failures" version of the negative binomial distribution. This version is related to the binomial distribution in a complementary fashion. That is, the binomial distribution can be used to model the number of successes $X$ we'd expect to see for a scenario where there are a fixed number of trials $N$ and a fixed probability of success $p$. In contrast, a negative binomial distribution can be used to model the number of failures $Y$ we'd expect to see before we encounter a fixed number of successes $r$ and a fixed probability of success $p$. 

In R, we can generate a random variable from a negative binomial distribution using this formulation as follows:

```{r tidy = F, warning = F, message = F}

set.seed(123)

y <- rnbinom(n = 5, prob = .2, size = 2) 

y

```

This tells us that we observed six failures before the first two successes, 16 failures before the next two successes, then eleven, then three, then seven. In other words, these represent the number of failures before we see two successes (since `size = 2`).

However, there is another way to formulate the negative binomial distribution, as a mixture of the Poisson and the Gamma distribution [@Casella2002; p196, exercise 4.34]. This formulation allows us to think of the negative binomial distribution as a Poisson distribution, but without the assumption that the variance of this distribution is exactly equal to the mean. This is the most common formulation used in epidemiologic settings [see, e.g., @Lloyd-Smith2005]. In R, this formulation can be deployed as:

```{r tidy = F, warning = F, message = F}

set.seed(123)

y <- rnbinom(n = 5, mu = 5, size = 2) 

y

```

In this formulation, the variance of the negative binomial random variable is defined as:
$$Var(Y) = \mu + \frac{\mu^2}{\text{size}},$$
so for small values of `size`, we generate data that has a variance greater than the mean. As we increase the `size` parameter, we decrease the variance. In the limit as size $\rightarrow \infty$, this negative binomial distribution becomes a Poisson distribution [@Cameron2013; p85].

<!-- This code shows that as size gets very large, nbinom becomes poisson. Importantly, the lamba in pois is the mu in binom. -->
<!-- ```{r tidy = F, warning = F, message = F} -->
<!-- set.seed(123) -->

<!-- y1 <- rnbinom(n = 50000, mu = 5, size = 1e7) -->

<!-- y2 <- rpois(n = 50000, lambda = 5) -->

<!-- hist(y1, breaks = 50, col = rgb(1, 0, 0, 0.5), main = "y1 vs y2", xlim = c(0, 20)) -->
<!-- hist(y2, breaks = 50, col = rgb(0, 0, 1, 0.5), add = TRUE) -->
<!-- legend("topright", legend = c("NB", "Poisson"), fill = c("red", "blue")) -->

<!-- ks.test(y1, y2) -->
<!-- ``` -->


:::{.rmdnote data-latex="{dive}"}

__Deeper Dive__: Overdispersion

|               Overdispersion describes a situation where the variance in an outcome variable exceeds the variance assumed under a model's distributional assumptions. This concept is most commonly encountered with a Poisson random variable. The variance of a Poisson distribution is equal to the mean. If we let $\mu$ denote the mean of an outcome variable $Y$, and we let $\sigma^2$ denote it's variance, then modeling $Y$ with a Poisson regression model (GLM or GEE) implies that $\mu = \sigma^2$. However, in actuality for some variables **assumed** to follow a Poisson distribution, we may find that: $\mu < \sigma^2$.

In this case, we would say that the outcome is "overdispersed". One solution to overdispersion is to estimate a scale parameter in a GLM or GEE model. In a Poisson regression context, adding a scale parameter re-defines the variance of the outcome to be:

$$V(Y) = \phi \mu,$$
where $\phi$ is a scale parameter that multiplies this variance accordingly. In a regression modeling context, we can set the scale parameter $\phi$ to 1 if there is no overdispersion. Or we can estimate it, and scale the standard errors from the regression model, thus accounting for overdispersion in the outcome.

Other distributions can be affected by overdispersion (e.g., binomial distribution with $N>1$). However, the Gaussian distribution and Binomial distribution with $N = 1$ (i.e., a Bernoulli random variable) are not typically affected by overdispersion. In these cases, it is common to set the scale parameter to 1 instead of estimate it. Other techniques are also available to handle overdispersion (e.g., using the Negative Binomal distribution instead, which allows the variance to exceed the mean for count outcome data).

:::

## Geometric Distribution

The geometric distribution models the number of trials one needs to get to a first success. It's called the "geometric" distribution because the probabilities that arise from this distribution follow a geometric sequence, which is straightforward to see if we examine the pdf of the distribution. 

```{r geomplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Histogram for the Geometric Distribution with p = 0.5 for 5000 Simulated Observations."}

set.seed(123)
ggplot() +
  geom_bar(aes(x = rgeom(n = 5000, prob = .5) )) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

In R, a geometric random variable can be generated as follows.

```{r tidy = F, warning = F, message = F}

set.seed(123)

y <- rgeom(n = 5, prob = .5)

y

```

This tells us that we had to try one time before we observed a success in the first round, then we had to try 3 times before observing a success, and then we achieved success on the first try for the next three rounds.

## Exponential Distribution

```{r exponentialplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Histogram for the Exponential Distribution with rate = 1 for 5000 Simulated Observations."}

set.seed(123)
ggplot() + 
  geom_histogram(aes(x = rexp(n = 5000, rate = 1)), bins = 50) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

The exponential distribution is often used in time-to-event analyses, where it models the time until a single event occurs. It is fully defined by a single parameter $\lambda$, the rate parameter. The mean of the exponential distribution is the inverse of the rate. The exponential distribution assumes a constant hazard over time, which is sometimes considered a limitation, and a reason that researchers rely on more flexible models, such as the Weibull or Generalized Gamma distributions (discussed later).

```{r tidy = F, warning = F, message = F}

set.seed(123)

n <- 5

y <- rexp(n, rate = 0.5)

y

```

In the absence of right censoring, time to event (or survival) outcomes are often characterized by two pieces of information: the time at which the event occurred, and an indicator of when the event occurred. The outcome `y` simulated from the exponential distribution with `rate = 0.5` does not have an event indicator. However, we can construct one using a few different motivations. 

The easiest way is to assume that all individuals experienced the event at the simulated time. For example, the first individual experienced the event at time 1.7, the second at time 1.2, the third at time 2.7, and so on. 

However, if we wanted to limit the proportion of events, we could set an arbitrary administrative censoring time. For instance, if we assume that our study followed individuals from time zero to time 1, we'd end up with the following time to event data:

```{r tidy = F, warning = F, message = F}

sim_dat <- data.frame(y_obs = ifelse(y<=1, y, 1), 
                      delta = as.numeric(y <= 1))

sim_dat

```

The `sim_dat` dataset shows us that the first three observations were followed from the start to the end of follow-up, at which point they were administratively censored, while the last two experienced the event during follow up. 

## Weibull Distribution

```{r weibullplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Histogram for the Weibull Distribution with a shape parameter = 2 and a scale parameter = 1 for 5000 Simulated Observations."}

set.seed(123)
ggplot() + 
  geom_histogram(aes(x = rweibull(n = 5000, shape = 2, scale = 1)), bins = 50) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

The Weibull distribution is a flexible distribution used to model survival data. It has two parameters: shape ($k$) and scale ($\lambda$). In R, we can use the `rweibull()` function to generate Weibull-distributed random variables.

```{r tidy = F, warning = F, message = F}

set.seed(123)
n <- 5
shape <- 2
scale <- 1
y <- rweibull(n, shape = shape, scale = scale)
y

```

The same ideas we discussed for the exponential distribution can be applied to the Weibull distribution. The key difference between them is that the Weibull need not assume a constant hazard over time. One can generate an exponentially distributed random variable from the `rweibull()` function as follows:

```{r tidy = F, warning = F, message = F}

set.seed(123)

n <- 5000
rate <- 2

y1 <- rweibull(n, shape = 1, scale = 1 / rate)

y2 <- rexp(n, rate = rate)

```

If we plot the `y1` and `y2` variables in a back to back histogram, we can see that the distribution is identical:

```{r tidy = F, warning = F, message = F, echo = F}
df <- data.frame(
  value = c(y2, y1),
  distribution = factor(rep(c("Exponential", "Weibull"), each = n))
)

# Create histogram bins for each distribution
bin_width <- 0.2
max_val <- max(df$value)
breaks <- seq(0, ceiling(max_val), by = bin_width)

hist_df <- df %>%
  mutate(bin = cut(value, breaks = breaks, include.lowest = TRUE, right = FALSE)) %>%
  group_by(distribution, bin) %>%
  summarise(count = n(), .groups = "drop") %>%
  mutate(
    count = ifelse(distribution == "Weibull", -count, count),
    bin_mid = as.numeric(sub("\\[(.+),.*", "\\1", bin)) + bin_width / 2
  )

# Plot: Vertical bars with central x-axis
p1 <- ggplot(hist_df, aes(x = bin_mid, y = count, fill = distribution)) +
  geom_col(width = bin_width, color = "black", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "black") +
  scale_y_continuous(labels = abs) +
  labs(
    x = "Random Variable Value",
    y = "Count"
  ) +
  theme(
    axis.title.y = element_text(margin = margin(r = 10)),
    axis.text.y = element_text(color = "black"),
    legend.title=element_blank()
  ) +
  scale_fill_manual(values = c("Exponential" = "steelblue", "Weibull" = "darkorange"))

ggsave(here("_images", "weibull_exponential_comparison.pdf"), width = 8, height = 8, units = "cm")

```

```{r expweib, out.width="8cm", fig.align='center', fig.cap="Comparison of the Distribution of a Random Variable Generated from the rexp() function with a rate parameter = 2, and the rweibull() function with shape = 1 and scale = 1 / rate.", echo=F}
knitr::include_graphics(here("_images", "weibull_exponential_comparison.pdf"))
```

This allows us to create an equivalence between the Weibull and exponential functions in R. The Weibull is an exponential when the shape parameter for the Weibull distribution is set to 1, and the scale parameter is set to the inverse of the rate of the exponential. 

This is important to note, since there are a number of different ways to parameterize the Weibull distribution, and it can get confusing if you don't know which one you're working with.

## Generalized Gamma Distribution

```{r gengammaplot, out.width="5cm", fig.align='center', fig.margin=TRUE, warning = F, message = F, echo=F, fig.cap="Histogram for the Generalized Gamma Distribution with a location parameter = log(2), a sigma parameter = 1, and a  Q parameter = 0.5 for 5000 Simulated Observations."}

pacman::p_load(flexsurv)
shape <- 2
scale <- 1
k <- 0.5
set.seed(123)
ggplot() + 
  geom_histogram(aes(x = rgengamma(n, mu = log(scale), sigma = 1/shape, Q = k)), bins = 50) +
  scale_x_continuous(expand = c(0,0)) +
  scale_y_continuous(expand = c(0,0))

```

The Generalized Gamma distribution is an even more flexible distribution used to model survival data. In terms of ranking, one might say that the exponential is the least flexible, followed by the Weibull, followed by the Generalized Gamma. Just as the Weibull distribution contains the exponential as a special case, the Generalized Gamma distribution contains the exponential and Weibull as special cases.

The Generalized Gamma distribution can simulate time-to-event outcomes with a wide range of hazard function shapes [@Cox2007], more than just the exponential (constant hazard) and Weibull can. It has three parameters: a location parameter denoted $\mu$, a scale parameter denoted $\sigma$, and a shape parameter denoted $Q$. In R, we can use the `rgengamma()` function in the `flexsurv` package to generate Generalized Gamma distributed random variables.

```{r tidy = F, warning = F, message = F}

pacman::p_load(flexsurv)

set.seed(123)
n <- 5
shape <- 2
scale <- 1
q <- 0.5

y <- rgengamma(n, mu = log(scale), sigma = 1/shape, Q = k)

y

```

To obtain an exponentially distributed random variable from this function, one should set both the `sigma` and `Q` parameters to one, and then set the `mu` parameter to be the log of the inverse of the rate parameter from an exponential distribution, as follows:

```{r}

set.seed(123)
n <- 5000
rate <- 2
scale <- 1 / rate

# Generate generalized gamma with exponential parameters
y <- rgengamma(n, mu = log(scale), sigma = 1, Q = 1)

# Compare with exponential
exp_y <- rexp(n, rate = rate)

# Kolmogorov-Smirnov test
ks.test(y, exp_y)


```


<!-- ## Mixture Distributions -->

<!-- ## Zero Inflated and Hurdle Distributions -->

<!-- ## Beta and Dirichlet -->

<!-- ## Hypergeometric -->
<!-- ## Chi Squared -->
<!-- ## Logistic -->
<!-- ## Gamma -->
<!-- ## Inverse Gaussian -->
<!-- ## Beta -->
<!-- ## Log Normal -->
<!-- ## Tweedie -->
<!-- ## Students t -->
<!-- ## Gompertz -->
<!-- ## Power Law Distributions -->
<!-- ## Pareto Type 1 and Type 2 (Lomax) -->

<!-- ## **ExtDist Package:** https://cran.r-project.org/web/packages/ExtDist/index.html -->

# The Inverse Transformation Method

Thus far, we've demonstrated a range of packages that can be used to simulate random variables from a wide variety of distributions. One strength of the R programming language is that it many functions that can be used for a diversity of tasks, including random variable generation.

But what happens if we want to simulate a random variable from a distribution, and there is no associated R package or function that would allow us to do this? Or what happens if we don't immediately trust the package that's available to us? For example, supposed I wanted to simulate from a Pareto distribution to generate a heavy-tailed random variable. I could use the corresponding suite of function from the `EnvStats` package, including the `rpareto()` functions, but let's say I don't trust the function because it is no longer being maintained, or that I suspect there might be an error in the function. Is there a way I can generate a random variable from a Pareto (or other) distribution that I can guarantee? 

It turns out, there is a generic mathematical tool we can use to do this. This tool is called the inverse transformation method, or inverse transform sampling. The approach allows us to derive an equation that we can write up in R, and then use a uniform random variable to draw random realizations from the distribution fo interest. 

Here's an example use of the inverse transform sampling method to generate data from a Pareto distribution with an 80:20 property. This distribution is useful to model some heavy tailed distributions. For example, consider the distribution of wealth in the United States. There are several lines of evidence suggesting that the distribution of wealth in the US is "heavy tailed." Some estimates demonstrate that the top 10% of individuals in the US possess 70% of the wealth, which is close to what we'd expect if wealth was distributed according to a Pareto random variable.

First, we note that the CDF of Pareto (type 1) distribution is defined as:

$$
F(x) = 
\begin{cases}
1 - \left( \dfrac{x_m}{x} \right)^{\alpha}, & \text{for } x \geq x_m \\\\
0, & \text{for } x < x_m
\end{cases}
$$

This equation defines the CDF for a Pareto distribution as a function of two parameters: $x_m$, which defines the minimum value the variable can take (the scale parameter); and $\alpha$, which defines the shape of the distribution. Since $F(x)$ is a cumulative probability, it lies on the [0,1] interval. So, we can let this distribution be equal to a draw from a uniform random variable on the [0,1] interval, and focus on the core part of this equation to get:

$$u = 1 - \left( \dfrac{x_m}{x} \right)^{\alpha}$$
With some basic algebra, we can re-arrange this equation:

$$x = x_m (1 - u)^{-1/\alpha}$$

This latter equation tells us that a realization of the random variable $x$ drawn from a Pareto distribution can be obtained from a draw from the uniform distribution $u$, if the uniform draw is entered into this equation. We can write R code that does this as:

```{r tidy = F, warning = F, message = F}

set.seed(123)

pareto_scale <- 10000 # this sets the minimum to the distribution
pareto_shape <- log(5, base = 4) # this sets the distribution to be Pareto 80:20

# now we use the inverse transform sampling method 
## generate uniform random variable
u <- runif(n = 1000)

## the equation that let's us apply inverse transform sampling from a uniform
## distribution to get a Pareto random variable
x <- pareto_scale*(1-u)^(-1/pareto_shape)

## Draw RV from same distribution but with the EnvStats Package function
## to compare the two random variables
pacman::p_load(EnvStats)
x_prime <- rpareto(n = 1000, 
                   location = pareto_scale, 
                   shape = pareto_shape)

```

We can compare the `x` and `x_prime` variables to see if our function worked. But because we generated from a Pareto distribution with a scale parameter of roughly $1.16 \approx \log_4(5)$, standard comparisons won't work very well. Figure \@ref(fig:paretocomparison) shows a log-log complementary CDF curve, that can be used to better understand how a heavy tailed random variable is behaving [@Clauset2009]. This Figure shows that both methods generate a random variables that have roughly the same tail distribution properties.

```{r paretocomparison, out.width="5cm", fig.align='center', fig.margin=TRUE, fig.cap="Log of the complement of the CDF against the log of X for the random variable generated using the inverse transform method (red lines) and the rpareto function in the EnvStats package (blue lines).", echo=F}
knitr::include_graphics(here("_images", "pareto_comparison.pdf"))
```

The inverse transform sampling method can be a powerful tool for generating random variables from a wide variety of distributions. The basic steps include:

1) Defining the CDF of the variable we want to generate as a uniform random variable
2) Invert the CDF so that the variable is isolated as a function of the distribution parameters and the uniform random variable
3) Program the equation in 2 using code
4) Simulate a uniform random variable bounded between 0 and 1, and enter it into the equation coded in 3

This inverse transform sampling method is a generic procedure that can be used for any distribution whose inverse CDF can be derived.

\newpage

# References