---
title: "Simulation Study: Example 1 - IP Weighting versus Marginal Standardization for a Time-Fixed Exposure"
author: "Ashley I Naimi"
date: "`r paste0('Spring ', 2025)`" #format(Sys.Date(), '%Y')
urlcolor: blue
bibliography: ../misc/ref.bib
link-citations: yes
output: 
    bookdown::pdf_book:
      base_format: tint::tintPdf
      toc: true
      number_sections: true
      includes:
        in_header: ../misc/preamble.tex
      latex_engine: xelatex
    html_document:
      theme: readable
      toc: true
      toc_float: true
      number_sections: true
      css: ../misc/style.css
---

```{r setup, include=FALSE}
library(knitr)
library(formatR)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

packages <- c( "data.table","tidyverse","ggplot2","ggExtra","formatR",
               "gridExtra","skimr","here","RColorBrewer","survival")

for (package in packages) {
  if (!require(package, character.only=T, quietly=T)) {
    install.packages(package, repos='http://lib.stat.cmu.edu/R/CRAN')
  }
}

for (package in packages) {
  library(package, character.only=T)
}

library(fontawesome)

thm <- theme_classic() +
  theme(
    legend.position = "top",
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )
theme_set(thm)

knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_chunk$set(echo = TRUE)

```

\newpage
\onehalfspacing

# Evaluating the Performance of IPW versus Marginal Standardization

We are going to wrap up our short course by pulling several of the pieces together that we've covered over the last few days. We'll use the ADEMP framework to develop a simulation study we'll conduct here. 

We'll start with the aims of our simulation study: to compare the performance of inverse probability weighting versus marginally standardized estimates of the average treatment effect. However, it would help us to get more specific with our intentions. For instance, we might ask whether we should choose IP-weighting versus marginal standardization with a different number of confounding variables (e.g., 10 versus 25. We might as whether we should choose the bootstrap versus the robust variance estimator when we use IP weighting. Getting more specific with our aims will help us make decisions about how we should construct our data generating mechanisms, analyses, and performance measures. 

# Aims

To start, let's specify some aims of our simulation study. These will help guide how we write the code we need to execute the tasks to fulfill our aims:

- Compare the performance of IP weighting versus marginal standardization when the number of confounding variables is 10 versus 25
- Evaluate the performance of the robust variance estimator versus the  bootstrap for the IP weighted estimator

We'll begin with these aims, but as we write our functions, it will be useful to keep in mind other variables that we may be interested in exploring. For example: we may want to look at what happens when the correlation between confounding variables changes; we may want to look at the impact of sample size; etc. We can easily incorporate these as arguments into our simulation function, even though we may decide to not explore them for the sake of, e.g., run time.

# Data Generating Mechanisms

In terms of generating our data, we will rely on a simple triangle DAG, with a set of confounding variables, a single exposure, and a single outcome:

```{r, out.width = "10cm",fig.cap="Simple confounding triangle, with exposure $X$, confounder $C$, and outcome $Y$.",echo=F}
knitr::include_graphics(here("_images","triangle_dag.pdf"))
```

This gives us a relatively straightforward causal ordering, as follows:

- The variable $C$ is most exogenous to this system
- The variable $X$ is second most exogenous
- The variable $Y$ is the most endogenous

Importantly, even though we may have a large number of variables $C$, because they are all exogenous and independent, we can treat them as a single node on the DAG, and simulate them simultaneously (and independently) in the code we write. 

# Estimands

We will focus our attention on the marginally adjusted odds ratio among the exposed versus unexposed:

$$\frac{odds(Y = 1 \mid X = 1)}{odds(Y = 1 \mid X = 0)}$$

where $odds(\bullet) = P(\bullet)/[1 - P(\bullet)]$. Note that, for a binary outcome, we could have also expressed interest in the marginally adjusted risk ratio, or the marginally adjusted risk difference, or some combination of all three. Since the code we will write for this section is already complex enough, we will forego targeting these estimands, but will note when and what changes can be made as we proceed with this example.

# Methods

## Inverse Probability Weighting

We're interested in comparing IP weighting to marginal standardization. Both of these estimators were introduced in Section 2, where we walked through their implementation step by step; here we restate only the essentials, and focus on what this example adds. To implement IP weighting, we need to first construct stabilized inverse probability weights, defined as:

\[
sw =
\begin{dcases}
\frac{P(X = 1)}{P(X = 1 \mid C)} & \text{if $X = 1$} \\
\frac{P(X = 0)}{P(X = 0 \mid C)} & \text{if $X = 0$}
\end{dcases}
\]

These are the same weights we constructed in Section 2. The only difference is notational: there, a single confounder $Z$ appeared in the denominator's conditioning set, while here we condition on the full set of confounders $C$. As before, the numerator can be estimated by taking the sample mean of $X$, and the denominator by taking fitted values from a logistic regression of the exposure against the confounders (i.e., the propensity score).

What is new in this example is the number of confounding variables. With 10 or 25 variables in $C$, writing out the propensity score model formula by hand is tedious, and complicates the evaluation of our methods when we change the number of confounders:

```{r, eval = F}

p_model <- glm(x ~ c1 + c2 + c3 + c4 + c5 + c6 + c7 + c8 + c9 + c10, data = the_data, family = binomial(link = "logit"))

```

To deal with this problem, we can use a coding trick in R that looks like this:

```{r, eval = F}

p_model <- glm(x ~ ., data = the_data, family = binomial(link = "logit"))

```

What the code in the above model does is it regresses the `x` variable against everything else in the dataset named `the_data`. This shorthand notation makes it easy to flexibly change what goes into a regression model. However, *we must ensure* that all the other variables in the dataset named `the_data` are those that we want in our model. If this is not the case, we will end up fitting a model that we do not want.

Inverse probability weighting can perform very poorly when the numerical value of the propensity score is very high or very low. For example, for an exposed individual, if we have an overall $P(X = 1) = 0.5$ (the numerator in our weight equation), and a propensity score of 0.00001, this would return a weight of:

$$0.5 / 0.00001 = 50000$$
This means that the individual with a propensity score of 0.00001 will contribute a total of 50,000 pseudo-individuals in our weighted analysis. If our original sample size was only 1,000, this would suggest an important problem: in the weighted analysis, a single individual is contributing 50,000 observations, which is 50 times the size of the actual data we have.

To address problems like this, we can trim the weights we create using the equation above. Specifically, we can construct a trimmed stabilized IP weight, where any values of the stabilized weights greater than (e.g.) the 97.5th percentile are **reset to the 97.5th percentile value**. For example, using an `ifelse` statement in R, we can construct trimmed weights as:

```{r, eval = F}

sw_trim <- ifelse(sw > quantile(sw, .975),
                  quantile(sw, .975),
                  analysis_data$sw)

```

## Marginal Standardization

Marginal standardization proceeds exactly as it did in Section 2: fit an outcome regression, generate predicted outcomes for everyone in the sample under the "exposed" and "unexposed" scenarios, and average each set of predictions. To implement this with a large number of confounding variables, we can use the same coding trick above to fit the outcome regression model as follows:

```{r, eval = F}

mu_model <- glm(y ~ ., data = the_data, family = binomial(link = "logit"))

```

For this modeling approach to work, we must ensure that the dataset used to fit this model (`the_data`) has **only** the outcome, the exposure, and the $p$ confounders we'd like to include in our model. All other variables, including ID variables, the estimated propensity score, the stabilized weights we might construct, and anything else, is excluded from the dataset used to fit this model. 

Once estimated, we can then generate predictions from this `mu_model` using the predict function, and averaging the predictions we obtained so that we can get our estimand:

```{r, eval = F}

muhat1 <- mean(predict(mu_model, newdata = transform(the_data, x=1), type="response"))

```

As in Section 2, `transform(the_data, x=1)` sets the exposure to 1 for every individual in the dataset, and `type="response"` returns the predictions on the probability scale rather than the log-odds scale. The same call with `x=0` gives us the unexposed counterpart, and averaging each set of predictions gives us the two marginal probabilities that make up our estimand.

# Performance Measures

Once we obtain our simulation data, we will use the following performance measures to compare IP weighting to marginal standardization:

- bias of the point estimate
- bias of the standard error
- confidence interval coverage
- confidence interval length

# Monte Carlo Integration for the True Marginal OR

In order to obtain estimates of our desired performance measures, we will need to compute the true marginal odds ratios for all the scenarios we will explore. In our case, we will limit our simulation to scenarios where the sample size for each Monte Carlo simulation is 500 and 1000, and where the number of confounders in each Monte Carlo simulation is 10 and 25. 

This means that we'll have to use Monte Carlo integration to estimate the true marginal odds ratios under *two* settings: when there are 10 and 25 confounders in the data generating mechanism.^[Note that we don't need to account for changing sample sizes in the Monte Carlo integration approach, since we are interested in maximizing the sample size for this step.] We can do this using the following code:

```{r tidy = F, eval = F}

 expit<-function(a){1/(1+exp(-a))}

 set.seed(123)
 simulation_function_true <- function(intercept = -2, 
                                      exposure = 1, 
                                      c_number = 10,
                                      cov_mat = 0,
                                      diag_cov = 1){
     n = 5e6

     # how many confounders?
     p <- c_number
     
     ## confounder matrix
     sigma <- matrix(cov_mat, nrow=p, ncol=p)
     diag(sigma) <- diag_cov
     c <- mvtnorm::rmvnorm(n, mean=rep(0,p), sigma=sigma)
     
     # DESIGN MATRIX FOR THE PROPENSITY SCORE MODEL
     piMat <- model.matrix(
       as.formula(
         paste("~(",
               paste("c[,",1:ncol(c),"]", collapse="+"),
               ")"
         )
       )
     )[,-1]
     
     # parameters for confounder exposure relation
     parmsC_pi <- rep(log(1.5), c_number)
     
     # simulate the exposure
     x <- exposure #rbinom(n, size = 1, expit(-.5 + piMat%*%parmsC_pi))
     
     # parameters for the confounder outcome relation
     parmsC_mu <- rep(log(2), c_number)
     
     # simulate the outcome
     pY <- mean(expit(intercept + log(2)*x + piMat%*%parmsC_mu))

     print(paste("the mean of y is: ", pY))

     ## what would we change here if interested in marginally adjusted risk difference or risk ratio?
     res <- pY/(1 - pY)

     return(res)
 }

 odds1 <- simulation_function_true(intercept = -2, exposure = 1, c_number = 10)
 odds0 <- simulation_function_true(intercept = -2, exposure = 0, c_number = 10)

 or_marg1 <- log(odds1/odds0)

 or_marg1

 odds1 <- simulation_function_true(intercept = -2, exposure = 1, c_number = 25)
 odds0 <- simulation_function_true(intercept = -2, exposure = 0, c_number = 25)

 or_marg2 <- log(odds1/odds0)

 or_marg2

 true_log_or1 <- or_marg1
 true_log_or2 <- or_marg2
 
 # the true log OR marg1 (c_number = 10) is: 0.4132932
 #
 # the true log OR marg2 (c_number = 25) is: 0.289803
```

# Running the Simulation

Now that we have our true values, we can go ahead and proceed with running the simulation. 

```{r tidy = F, warning = F, message = F, eval = F}

 library(parallel)
 library(lmtest)
 library(sandwich)

 expit<-function(a){1/(1+exp(-a))}

 set.seed(123)

 simulation_function <- function(index,
                                 intercept=-2,
                                 sample_size = 1000,
                                 c_number = 10,
                                 cov_mat = 0,
                                 diag_cov = 1){

     # printing to console won't work with parallel processing
     print(index)

     # DATA GENERATION
     n <- sample_size
     print(n)

     # how many confounders to simulate?
     p <- c_number
     print(p)

     ## confounder matrix
     sigma <- matrix(cov_mat, nrow=p, ncol=p)
     diag(sigma) <- diag_cov
     c <- mvtnorm::rmvnorm(n, mean=rep(0,p), sigma=sigma)

     # DESIGN MATRIX FOR THE PROPENSITY SCORE MODEL
     piMat <- model.matrix(
       as.formula(
         paste("~(",
               paste("c[,",1:ncol(c),"]", collapse="+"),
               ")"
               )
         )
       )[,-1]

     # parameters for confounder exposure relation
     parmsC_pi <- rep(log(1.5), c_number)

     # simulate the exposure
     x <- rbinom(n, size = 1, expit(-.5 + piMat%*%parmsC_pi))

     # parameters for the confounder outcome relation
     parmsC_mu <- rep(log(2), c_number)

     # simulate the outcome
     y <- rbinom(n, size = 1, expit(intercept + log(2)*x + piMat%*%parmsC_mu))

     # construct dataset
     analysis_data <- data.frame(y, x, c)

     # ANALYSIS

     # marginal standardization
     meanY <- mean(analysis_data$y)

     m1 <- glm(y ~ ., data = analysis_data, family=binomial(link="logit"))

     muhat1 <- mean(predict(m1, newdata = transform(analysis_data, x = 1), type="response"))
     muhat0 <- mean(predict(m1, newdata = transform(analysis_data, x = 0), type="response"))

     ## compute the odds from these average probabilities
     odds1<-muhat1/(1-muhat1)
     odds0<-muhat0/(1-muhat0)

     ## marginal log-odds ratio
     marginal_standardization_estimate <- log(odds1/odds0)

     # IP WEIGHTING
     # create the propensity score in the dataset
     analysis_data$propensity_score <- glm(x ~ .,
                                           data = analysis_data %>% select(-y),
                                           family = binomial("logit"))$fitted.values

     # stabilized inverse probability weights
     analysis_data$sw <- (mean(analysis_data$x)/analysis_data$propensity_score)*analysis_data$x +
       ((1-mean(analysis_data$x))/(1-analysis_data$propensity_score))*(1-analysis_data$x)
     
     summary(analysis_data$sw)
     quantile(analysis_data$sw, .995)
     
     # trim the weights
     analysis_data$sw <- ifelse(analysis_data$sw>quantile(analysis_data$sw, .995),
                                quantile(analysis_data$sw, .995),
                                analysis_data$sw)
     
     # outcome model
     m2 <- glm(y ~ x, data = analysis_data, weights = sw, family = quasibinomial(link="logit"))

     ip_weighting_estimate <- coeftest(m2,
                                       vcov. = vcovHC(m2, type = "HC3"))[2,1:2]

     # bootstrap SEs
     boot_func <- function(nboot, data){

       a_dat <- data

       index <- sample(1:nrow(a_dat), nrow(a_dat), replace = T)

       boot_dat <- a_dat[index, ]

       # MARGINAL STANDARDIZATION
       m1_ <- glm(y ~ ., data = boot_dat, family=binomial(link="logit"))

       muhat1_ <- mean(predict(m1_, newdata = transform(boot_dat, x=1), type="response"))
       muhat0_ <- mean(predict(m1_, newdata = transform(boot_dat, x=0), type="response"))

       ## compute the odds from these average probabilities
       odds1_ <- muhat1_/(1-muhat1_)
       odds0_ <- muhat0_/(1-muhat0_)

       marginal_stand_ <- log(odds1_/odds0_)

       # IP WEIGHTING
       # create the propensity score in the dataset
       boot_dat$propensity_score <- glm(x ~ .,
                                        data = boot_dat %>% select(-y),
                                        family = binomial("logit"))$fitted.values

       # stabilized inverse probability weights
       boot_dat$sw <- (mean(boot_dat$x)/boot_dat$propensity_score)*boot_dat$x +
         ((1-mean(boot_dat$x))/(1-boot_dat$propensity_score))*(1-boot_dat$x)
       
       # trim the weights
       boot_dat$sw <- ifelse(boot_dat$sw>quantile(boot_dat$sw, .995),
                             quantile(boot_dat$sw, .995),
                             boot_dat$sw)

       # outcome model
       m2_ <- glm(y ~ x,
                  data = boot_dat,
                  weights = sw,
                  family = quasibinomial(link="logit"))

       ip_weighting_ <- summary(m2_)$coefficients[2,1]

       return(c(marginal_stand_, ip_weighting_))
     }

     analysis_data <- analysis_data %>% select(-propensity_score, -sw)
     
     boot_res <- lapply(1:500, function(x) boot_func(x, data = analysis_data))

     boot_SE <- apply(do.call(rbind, boot_res), 2, sd)

     # SIMULATION FUNCTION OUTPUT

     res <- data.frame(intercept = intercept,
                       sample_size = sample_size,
                       c_number = c_number,
                       cov_mat  = cov_mat,
                       diag_cov = diag_cov,
                       marginal_standardization_estimate,
                       marginal_standardization_SE = boot_SE[1],
                       ip_weighting_estimate = ip_weighting_estimate[1],
                       ip_weighting_robust_SE = ip_weighting_estimate[2],
                       ip_weighting_boot_SE = boot_SE[2])

     return(res)
 }

 # simulation function parameters

 parm_data <- expand.grid(
   index = 1:200,
   sample_size = 1000,
   intercept = -2,
   c_number = c(10, 25),
   cov_mat = 0,
   diag_cov = 1
 )

 head(parm_data, 3)

 tail(parm_data, 3)

 simulation_results <- mclapply(1:nrow(parm_data),
                              function(x) simulation_function(index = parm_data[x,]$index,
                                                              intercept = parm_data[x,]$intercept,
                                                              sample_size = parm_data[x,]$sample_size,
                                                              c_number = parm_data[x,]$c_number,
                                                              cov_mat = parm_data[x,]$cov_mat,
                                                              diag_cov = parm_data[x,]$diag_cov),
                              mc.cores = detectCores() - 2)

 sim_res <- do.call(rbind, simulation_results)

 # # save the data to file!

 head(sim_res)
 
 write_csv(sim_res, here("lectures/06_Section6_SimulationInPractice", "simulation_results_MSIPW.csv"))

```

# Performance Evaluation

This simulation takes some time to run, so let's be sure we save our results to a file so that we can re-import them without having to run the simulation again to rebuild the entire dataset. We can import our saved data back into R using standard code, such as the `read` functions in the tidyverse:

```{r tidy = F, warning = F, message = F}

a <- read_csv(here("lectures/06_Section6_SimulationInPractice", "simulation_results_MSIPW.csv"))

head(a)

# to make things easier, let's create objects with our true values:

true_log_or1 <- 0.4132932 # true log OR when c_number = 10
true_log_or2 <- 0.289803  # true log OR when c_number = 25

# the true value depends on the scenario (number of confounders),
# not on the estimator, so we attach the appropriate truth to each row:

a <- a %>%
  mutate(true_log_or = ifelse(c_number == 10, true_log_or1, true_log_or2))

```

Let's start with the mean of the point estimates obtained from marginal standardization and IP weighting:

```{r tidy = F, warning = F, message = F}

# the true log OR marg1 (c_number = 10) is: 0.4132932
#
# the true log OR marg2 (c_number = 25) is: 0.289803

a %>% 
  group_by(c_number) %>% 
  summarize(meanMS = mean(marginal_standardization_estimate),
            meanIPW = mean(ip_weighting_estimate))

```

These results suggest some potential problems with the IP weighting approach. Let's compute the bias:

```{r tidy = F, warning = F, message = F}

# the true log OR marg1 (c_number = 10) is: 0.4132932
#
# the true log OR marg2 (c_number = 25) is: 0.289803

a %>%
  group_by(c_number) %>%
  summarize(biasMS = mean(marginal_standardization_estimate - true_log_or),
            biasIPW = mean(ip_weighting_estimate - true_log_or))

```

The bias estimates for IP weighting seem fairly large, while those for marginal standardization are much smaller. We can ask if these estimates are compatible with an underlying hypothesis of "no bias", after all, we only did 200 simulations, which is a fairly low number. Let's compute standard errors for each of these bias estimates. Also, it might help us to construct a dataset with all these results:

```{r tidy = F, warning = F, message = F}
mc_se_bias <- function(x, n){
  sqrt(sum((x - mean(x))^2)/(n*(n-1)))
}

bias_results <- a %>%
  group_by(c_number) %>%
  summarize(biasMS = mean(marginal_standardization_estimate - true_log_or),
            biasIPW = mean(ip_weighting_estimate - true_log_or),

            biasMS_se = mc_se_bias(marginal_standardization_estimate - true_log_or, n = 200),
            biasIPW_se = mc_se_bias(ip_weighting_estimate - true_log_or, n = 200),
            
            biasMS_p.value = round(2*(1 - pnorm(abs(biasMS/biasMS_se))),4),
            biasIPW_p.value = round(2*(1 - pnorm(abs(biasIPW/biasIPW_se))),4))

bias_results

```


```{r tidy = F, warning = F, message = F}

ggplot(a) +
  geom_histogram(aes(marginal_standardization_estimate - true_log_or)) +
  geom_vline(xintercept = 0, color = "red") +
  facet_wrap(~c_number)

```


```{r tidy = F, warning = F, message = F}

ggplot(a) +
  geom_histogram(aes(ip_weighting_estimate - true_log_or)) +
  geom_vline(xintercept = 0, color = "red") +
  facet_wrap(~c_number)

```

## Bias of the Standard Errors

Bias of the point estimates addresses our first aim, but it says nothing yet about our second aim: how well the robust variance estimator and the bootstrap estimate the uncertainty in the IP weighted estimator. Recall from Section 4 that a standard error is meant to estimate the standard deviation of the point estimates across repeated samples. Our Monte Carlo simulation gives us exactly that repeated-sampling distribution, so we can compare the average of the estimated standard errors in each scenario to the empirical standard deviation of the point estimates in that scenario. Our simulation output contains three estimated standard errors: the bootstrap standard error for marginal standardization, and both the robust (sandwich) and bootstrap standard errors for IP weighting:

```{r tidy = F, warning = F, message = F}

se_results <- a %>%
  group_by(c_number) %>%
  summarize(empSE_MS = sd(marginal_standardization_estimate),
            meanSE_MS = mean(marginal_standardization_SE),

            empSE_IPW = sd(ip_weighting_estimate),
            meanSE_IPW_robust = mean(ip_weighting_robust_SE),
            meanSE_IPW_boot = mean(ip_weighting_boot_SE),

            ratio_MS = empSE_MS/meanSE_MS,
            ratio_IPW_robust = empSE_IPW/meanSE_IPW_robust,
            ratio_IPW_boot = empSE_IPW/meanSE_IPW_boot)

se_results

```

The ratios in this table divide the empirical (true) standard error by the average estimated standard error, so a value of 1 indicates a perfectly calibrated standard error estimator, a value below 1 indicates a standard error that is too large (conservative), and a value above 1 indicates a standard error that is too small (anti-conservative). The bootstrap standard error for marginal standardization is well calibrated in both scenarios (ratios of roughly 0.98 and 1.02). For IP weighting, the two variance estimators behave quite differently, which speaks directly to our second aim. The bootstrap standard error tracks the true sampling variability closely (ratios of roughly 0.95 in both scenarios). The robust standard error, by contrast, overestimates the true sampling variability by roughly 30% (ratios of roughly 0.75). This conservatism is a known property of the robust variance estimator in this setting: it treats the estimated weights as if they were known, and ignoring the estimation of the weights leads it to overstate the variability of the stabilized IP weighted estimator.

## Confidence Interval Coverage

Point estimates and standard errors come together in confidence interval coverage: the proportion of Monte Carlo samples in which the 95% Wald interval contains the true value. As in Section 4, we construct the intervals as the estimate plus or minus 1.96 standard errors, which gives us three intervals to evaluate — one for marginal standardization, and one for each of the two IP weighting variance estimators:

```{r tidy = F, warning = F, message = F}

a <- a %>%
  mutate(MS_LCL = marginal_standardization_estimate - 1.96*marginal_standardization_SE,
         MS_UCL = marginal_standardization_estimate + 1.96*marginal_standardization_SE,

         IPW_robust_LCL = ip_weighting_estimate - 1.96*ip_weighting_robust_SE,
         IPW_robust_UCL = ip_weighting_estimate + 1.96*ip_weighting_robust_SE,

         IPW_boot_LCL = ip_weighting_estimate - 1.96*ip_weighting_boot_SE,
         IPW_boot_UCL = ip_weighting_estimate + 1.96*ip_weighting_boot_SE)

coverage_results <- a %>%
  group_by(c_number) %>%
  summarize(coverageMS = mean(MS_LCL < true_log_or & true_log_or < MS_UCL),
            coverageIPW_robust = mean(IPW_robust_LCL < true_log_or & true_log_or < IPW_robust_UCL),
            coverageIPW_boot = mean(IPW_boot_LCL < true_log_or & true_log_or < IPW_boot_UCL))

coverage_results

```

Before interpreting these numbers, we should ask how much Monte Carlo error to expect. Coverage is a proportion, so its Monte Carlo standard error is $\sqrt{p(1-p)/n_{sim}}$ [@Morris2019]. With 200 simulations and a nominal coverage of 0.95, that is:

```{r tidy = F, warning = F, message = F}

sqrt(.95*(1 - .95)/200)

```

So with only 200 simulations, estimated coverage within roughly 1.5 to 3 percentage points of 95% is compatible with nominal coverage. Marginal standardization covers at 96.5% with 10 confounders and 92% with 25 confounders — the first clearly compatible with the nominal rate, the second borderline (about two Monte Carlo standard errors low). The IP weighting intervals tell a more interesting story. With 10 confounders, the robust interval covers at 97.5% — above nominal, because the conservative robust standard error produces intervals that are wider than they need to be, which also happens to absorb the modest bias in the IP weighted point estimate. The bootstrap interval, built from a well-calibrated standard error, covers at 93.5%. With 25 confounders, however, both IP weighting intervals fall well short of nominal: 88.5% for the robust interval and 77% for the bootstrap interval. This is the bias from our first aim coming home to roost: when the point estimate is centered in the wrong place, even a perfectly estimated standard error yields intervals that miss the truth too often. Indeed, the well-calibrated bootstrap interval covers *worse* than the conservative robust interval here, because the robust interval's excess width partially compensates for the bias. This is why it is worth evaluating standard error bias and coverage separately: coverage mixes together the quality of the point estimator and the quality of the variance estimator, and only by looking at both measures can we tell which one is responsible for a coverage problem.

## Confidence Interval Length

Coverage alone can be misleading — we can always achieve high coverage with absurdly wide intervals. So our final performance measure is the average length of the confidence intervals:

```{r tidy = F, warning = F, message = F}

length_results <- a %>%
  group_by(c_number) %>%
  summarize(lengthMS = mean(MS_UCL - MS_LCL),
            lengthIPW_robust = mean(IPW_robust_UCL - IPW_robust_LCL),
            lengthIPW_boot = mean(IPW_boot_UCL - IPW_boot_LCL))

length_results

```

The marginal standardization intervals are the shortest in both scenarios, and they get *shorter* as the number of confounders grows (from 0.48 to 0.41) — conditioning on more outcome predictors improves the precision of the standardized estimate. The IP weighting intervals move in the opposite direction, growing with the number of confounders, and the robust intervals are the longest of all: roughly 46% longer than the marginal standardization intervals with 10 confounders, and more than twice as long with 25.

# Summarizing the Results

We can now answer the two aims we set out at the beginning of this simulation study. For our first aim, marginal standardization clearly outperforms IP weighting in these scenarios: it is essentially unbiased with both 10 and 25 confounders, its intervals are the shortest, and its coverage is at or near the nominal rate. IP weighting shows bias that grows with the number of confounders, and with 25 confounders neither of its interval estimators achieves close to nominal coverage.

For our second aim, the bootstrap estimates the true sampling variability of the IP weighted estimator well, while the robust variance estimator overestimates it by roughly 30%, yielding intervals that are conservative by construction. If we had to choose a variance estimator for IP weighting in settings like these, the bootstrap is the more accurate one — its standard errors mean what they claim to mean. The higher coverage of the robust intervals in these scenarios is not a point in their favor: it comes from excess width that happens to mask the estimator's bias, which is a property we should fix in the estimator (for example, with a better propensity score model specification or a different trimming strategy), not one we should paper over with an inflated variance.

\newpage

# References