---
title: "Simulation Study: Example 3 - Causal Mediation Analysis"
author: "Ashley I Naimi"
date: "`r paste0('Spring ', 2025)`" #format(Sys.Date(), '%Y')
urlcolor: blue
bibliography: ref.bib
link-citations: yes
output: 
    bookdown::pdf_book:
      base_format: tint::tintPdf
      toc: true
      number_sections: true
      includes:
        in_header: ../misc/preamble.tex
      latex_engine: xelatex
    html_document:
      theme: readable
      toc: true
      toc_float: true
      number_sections: true
      css: ../misc/style.css
---

```{r setup, include=FALSE}
library(knitr)
library(formatR)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)

packages <- c( "data.table","tidyverse","ggplot2","ggExtra","formatR",
               "gridExtra","skimr","here","RColorBrewer","survival")

for (package in packages) {
  if (!require(package, character.only=T, quietly=T)) {
    install.packages(package, repos='http://lib.stat.cmu.edu/R/CRAN')
  }
}

for (package in packages) {
  library(package, character.only=T)
}

remotes::install_github("rstudio/fontawesome")

library(fontawesome)

thm <- theme_classic() +
  theme(
    legend.position = "top",
    legend.background = element_rect(fill = "transparent", colour = NA),
    legend.key = element_rect(fill = "transparent", colour = NA)
  )
theme_set(thm)

knitr::knit_hooks$set(purl = knitr::hook_purl)
knitr::opts_chunk$set(echo = TRUE)

```

\newpage
\onehalfspacing

# Estimating the Controlled Direct Effect

Our final example of using simulation will involve causal mediation analysis. We are going to use the data generating mechanism we used that was based on the following DAG:

```{r, out.width = "10cm",fig.cap="Causal diagram representing exposure induced mediator outcome confounding.",echo=F}
knitr::include_graphics(here("_images","mediation_dag.pdf"))
```

Data generated from this mechanism can be used to quantify the controlled direct effect, defined as:

$$E(Y^{x = 1, z} - Y^{x = 0, z})$$

# Aims

In terms of aims, let's evaluate what happens when we improperly adjust for the confounder of the mediator outcome association that is downstream to the exposure. There are a few ways in which we can improperly adjust for this confounder, and we'll explore two:

- Evaluate the bias of a CDE estimator when we include the exposure induced mediator outcome confounder ($L$) in a simple regression model
- Evaluate the bias of a CDE estimator when we fail to adjust for the exposure induced mediator outcome confounder

As we saw previously, the DAG above suggests the following causal ordering:

|  Variable  | Arrow Number |
|----|---|
| $U$  |  0 |
| $C$  |  0 |
| $X$  |  1 |
| $L$  |  2 |
| $Z$  |  2 |
| $Y$  |  5 |

# Monte Carlo Integration for the True Value

In order to evaluate the bias, we're going to need to compute the true value. We can do this here using Monte Carlo integration. Let's write a program that adapts our data generating mechanism so that we can compute the true CDM in the setting we'd like to explore:

```{r tidy = F, warning = F, message = F}

expit <- function(x){1/(1 + exp(-x))}

set.seed(123)

true_CDM <- function(sample_size, exposure, mediator){
  
  n <- sample_size

  U <- rnorm(n)
  
  C <- rnorm(n)
  
  X <- exposure # rbinom(n, size = 1, p = expit(-1 + log(2)*C ))
  
  L <- rnorm(n, mean = 0 + 1.5*X + 1.5*U)
  
  Z <- mediator # rbinom(n, size = 1, expit(-1.5 + log(2)*X + log(2)*L ))
  
  return(mean(10 + 5*X + 5*C + 5*Z + 4*L + 4*U))

}


```


# Estimands




\newpage

# References